---
title: "Homework 02"
author: "Alec L. Robitaille"
date: "2024-02-15"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> From the Howell1 dataset, consider only the people younger than 13 years old.
Estimate the causal association between age and weight. Assume that age
influences weight through two paths. First, age influences height, and height
influences weight. Second, age directly influences weight through age- related
changes in muscle growth and body proportions.

> Draw the DAG that represents these causal relationships. And then write a
generative simulation that takes age as an input and simulates height and
weight, obeying the relationships in the DAG.


```{r}
coords <- data.frame(
    name = c('H', 'A', 'W'),
    x =    c(2,    1,   1),
    y =    c(1,    2,   0)
)

dagify(
    W ~ H + A,
    H ~ A,
  coords = coords
) |> ggdag(seed = 2) + theme_dag()
```


```{r}
n <- 1e2
max_age <- 13

simulate_hw <- function(n, max_age) {
	age <- runif(n, 0, max_age)
	
	# General vague height at 13 years old (cm) = 130
	beta_age_on_height <- 10
	height <- rnorm(n = n, mean = age * beta_age_on_height, 2)

	# General vague weight at 13 years old (kg) = 45
	# Betas chosen to hit 45 kg approximately at 13 years old
	beta_height_on_weight <- 0.2
	beta_age_on_weight <- 1.5
	weight <- (age * beta_age_on_weight) + 
		(beta_height_on_weight * height)
	
	data.table(age, height, weight)
}
ggplot(simulate_hw(n, max_age)) + 
	geom_point(aes(age, weight))
```

## Question 2

> Estimate the total causal effect of each year of growth on weight.

The total causal effect of age on weight is estimated by excluding the height
variable to include both paths in the DAG from age to weight directly and
indirectly through height ([Lecture 4: slide 28](https://speakerdeck.com/rmcelreath/statistical-rethinking-2023-lecture-04?slide=28)).


### Prior predictive simulation

```{r}
# Function for preparing Howell data
data_Howell
```

```{r}
# Load data
DT <- tar_read(Howell_lte_13)

# Print priors used
tar_read(h02_q02_brms_prior)

# Load model
tar_load(h02_q02_brms_sample_prior)
h02_q02_brms_sample_prior

# Read N draws from the priors and append expected predictions
n_draws <- 50
q02_draws <- h02_q02_brms_sample_prior |>
	add_epred_draws(newdata = unique(DT[, .(age)]),
									ndraws = n_draws)

# Plot prior expectations for relationship between age and weight
ggplot(q02_draws) + 
	geom_line(aes(age, .epred, group = .draw), alpha = 0.5) +
	labs(x = 'Age', y = 'Weight')
```

### Analyse the data

```{r}
# Load model
tar_load(h02_q02_brms_sample)
h02_q02_brms_sample

# Tidy draws
q02_draws <- h02_q02_brms_sample |> 
	tidy_draws()

q02_newdata <- data_grid(DT, age = seq_range(age, 20))
q02_epred <- h02_q02_brms_sample |> 
	epred_draws(q02_newdata)

# Total causal effect of each year of growth on weight
g1 <- ggplot(q02_draws) + 
	stat_halfeye(aes(b_age)) + 
	labs(x = 'Increase in weight per year of growth', y = '')

g2 <- ggplot(q02_epred) + 
	stat_ribbon(aes(age, .epred), alpha = 0.5) +
	scale_fill_grey(start = 0.8, end = 0.2) + 
	labs(x = 'Age', y = 'Weight')

g1 + g2
```

## Question 3

> The data in data(Oxboys) (rethinking package) are growth records for 26 boys
measured over 9 periods. I want you to model their growth. Specifically, model
the increments in growth from one period (Occasion in the data table) to the
next. Each increment is simply the difference between height in one occasion and
height in the previous occasion. Since none of these boys shrunk during the
study, all of the growth increments are greater than zero. Estimate the
posterior distribution of these increments. Constrain the distribution so it is
always positive—it should not be possible for the model to think that boys can
shrink from year to year. Finally compute the posterior distribution of the
total growth over all 9 occasions.


### Prior predictive simulation

```{r}
# Function for preparing Oxboys data
data_Oxboys
```

```{r}
# Load data
DT <- tar_read(prep_Oxboys)

# Print priors used
tar_read(h02_q03_brms_prior)

# Load model
tar_load(h02_q03_brms_sample_prior)
h02_q03_brms_sample_prior

# Read N draws from the priors and append expected predictions
n_draws <- 100
q03_newdata <- na.omit(DT)[, .(occasion_factor = unique(occasion_factor))]
q03_epred_prior <- h02_q03_brms_sample_prior |>
	add_epred_draws(newdata = q03_newdata, ndraws = n_draws)

# Plot prior expectations for relationship between occasion and height
ggplot(q03_epred_prior) + 
	stat_halfeye(aes(occasion_factor, .epred), alpha = 0.5) +
	labs(x = 'Occasion', y = 'Difference in height')
```
