---
title: "Homework 03"
author: "Alec L. Robitaille"
date: "2024-02-28"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> The first two problems are based on the same data. The data in data(foxes) are
116 foxes from 30 different urban groups in England. These fox groups are like
street gangs. Group size (groupsize) varies from 2 to 8 individuals. Each group
maintains its own (almost exclusive) urban territory. Some territories are
larger than others. The area variable encodes this information. Some territories
also have more avgfood than others. And food influences the weight of each fox.
Assume [the DAG below], where F is avgfood, G is groupsize, A is area, and W is
weight. Use the backdoor criterion and estimate the total causal influence of A
on F. What effect would increasing the area of a territory have on the amount of
food inside it?


### Estimand 

What is the total causal influence of A on F? What effect would increasing the
area of a territory have on the amount of food inside it?


### Scientific model

```{r}
#| echo: false
coords <- data.frame(
	name = c('F', 'G', 'A', 'W'),
	x =    c(1,    2,   1.5, 1.5),
	y =    c(0,    0,   1, -1)
)
```

```{r}
dag <- dagify(
	G ~ F,
	F ~ A,
	W ~ F + G,
  coords = coords,
	exposure = 'A',
	outcome = 'F'
)
ggdag_status(dag, seed = 2, layout = 'auto') + theme_dag()
```


#### Backdoor criterion

1. Identify all paths connecting treatment to the outcome, regardless of the
direction of arrows

- `A -> F`

2. Identify paths with arrows entering the treatment (backdoor). These are 
non-casual paths, because causal paths exit the treatment (frontdoor). 

- ~~`A -> F`~~


3. Find adjustment sets that close all backdoor/non-causal paths. 

There are no backdoor paths entering the treatment (A). The adjustment set is empty. 

```{r}
ggdag_adjustment_set(dag, effect = 'total') + theme_dag()
```


### Statistical model

$$F_{i} \sim Normal(\mu_{i}, \sigma)$$
$$\mu_{i} = \alpha + \beta_{A} A_{i}$$

$$\alpha \sim Normal(0, 0.2)$$

$$\beta_{A} \sim Normal(0, 0.5)$$

$$\sigma \sim Exponential(1)$$

#### Prior predictive simulation

```{r}
n <- 100
ggplot() + 
	geom_abline(aes(intercept = rnorm(n, 0, 0.2),
									slope = rnorm(n, 0, 0.5)),
							alpha = 0.1, linewidth = 2) + 
	labs(x = 'area (standardized)', 
			 y = 'average food (standardized)',
			 title = TeX(r"($\alpha \sim Normal(0, 0.2)$, $\beta_{F} \sim Normal(0, 0.5)$)")) + 
	xlim(-2, 2) + 
	ylim(-2, 2)
```


### Analyze the data

`?foxes` 

- group : ID of group
- avgfood : Average available food in group's territory
- groupsize : Size of each group
- area : Area of group territory
- weight : Body weight of individual fox

```{r}
data_foxes
```

```{r}
# Load data
DT <- tar_read(scaled_foxes)

# Print priors used
tar_read(h03_q01_brms_prior)

# Load model
tar_load(h03_q01_brms_sample)
h03_q01_brms_sample

# Tidy draws
q01_draws <- h03_q01_brms_sample |> 
    tidy_draws()

q01_newdata <- data_grid(DT, scale_area = seq_range(scale_area, 20))
q01_epred <- h03_q01_brms_sample |> 
    epred_draws(q01_newdata)

# Total causal influence of area on avg food
g1 <- ggplot(q01_epred) +
	stat_ribbon(aes(
		reverse_scale(scale_area, attributes(DT$scale_area)),
		reverse_scale(.epred, attributes(DT$scale_avgfood))
	), alpha = 0.5) +
	scale_fill_grey(start = 0.8, end = 0.2) +
	labs(x = 'Area', y = 'Average food')

g2 <- ggplot(q01_draws) +
	geom_vline(xintercept = 0, color = 'grey70') +
	stat_halfeye(aes(b_scale_area)) +
	labs(x = 'Total causal influence of area on average food', y = '') + 
	xlim(-1.2, 1.2)

g1 / g2
```


## Question 2


> Infer the total causal effect of adding food F to a territory on the weight W
of foxes. Can you calculate the causal effect by simulating an intervention on
food?


### Estimand 

What is the total causal influence of F on W? What effect would increasing the
amount of food have on weights?


### Scientific model

```{r}
#| echo: false
coords <- data.frame(
	name = c('F', 'G', 'A', 'W'),
	x =    c(1,    2,   1.5, 1.5),
	y =    c(0,    0,   1, -1)
)
```

```{r}
dag <- dagify(
	G ~ F,
	F ~ A,
	W ~ F + G,
  coords = coords,
	exposure = 'F',
	outcome = 'W'
)
ggdag_status(dag, seed = 2, layout = 'auto') + theme_dag()
```


#### Backdoor criterion

1. Identify all paths connecting treatment to the outcome, regardless of the
direction of arrows

- `F -> W`
- `F -> G -> W`

2. Identify paths with arrows entering the treatment (backdoor). These are 
non-casual paths, because causal paths exit the treatment (frontdoor). 

- ~~`F -> W`~~
- ~~`F -> G -> W`~~


3. Find adjustment sets that close all backdoor/non-causal paths. 

There are no backdoor paths entering the treatment (F). The adjustment set is empty. 

```{r}
ggdag_adjustment_set(dag, effect = 'total') + theme_dag()
```


### Statistical model

$$W_{i} \sim Normal(\mu_{i}, \sigma)$$
$$\mu_{i} = \alpha + \beta_{F} F_{i}$$

$$\alpha \sim Normal(0, 0.2)$$

$$\beta_{F} \sim Normal(0, 0.5)$$

$$\sigma \sim Exponential(1)$$

#### Prior predictive simulation

```{r}
n <- 100
ggplot() + 
	geom_abline(aes(intercept = rnorm(n, 0, 0.2),
									slope = rnorm(n, 0, 0.5)),
							alpha = 0.1, linewidth = 2) + 
	labs(x = 'average food (standardized)', 
			 y = 'weight (standardized)',
			 title = TeX(r"($\alpha \sim Normal(0, 0.2)$, $\beta_{F} \sim Normal(0, 0.5)$)")) + 
	xlim(-2, 2) + 
	ylim(-2, 2)
```


### Analyze the data

```{r}
# Print priors used
tar_read(h03_q02_brms_prior)

# Load model
tar_load(h03_q02_brms_sample)
h03_q02_brms_sample

# Tidy draws
q02_draws <- h03_q02_brms_sample |> 
    tidy_draws()

q02_newdata <- data_grid(DT, scale_avgfood = seq_range(scale_avgfood, 20))
q02_epred <- h03_q02_brms_sample |> 
    epred_draws(q02_newdata)

# Total causal influence of area on avg food
g1 <- ggplot(q02_epred) +
	stat_ribbon(aes(
		reverse_scale(scale_avgfood, attributes(DT$scale_avgfood)),
		reverse_scale(.epred, attributes(DT$scale_weight))
	), alpha = 0.5) +
	scale_fill_grey(start = 0.8, end = 0.2) +
	labs(x = 'Average food', y = 'Weight') +
	ylim(2, 6)

g2 <- ggplot(q02_draws) +
	geom_vline(xintercept = 0, color = 'grey70') +
	stat_halfeye(aes(b_scale_avgfood)) +
	labs(x = 'Total causal influence of average food on weight', y = '') + 
	xlim(-1.2, 1.2)

g1 / g2
```



## Question 3

> Infer the direct causal effect of adding food F to a territory on the weight W
of foxes. In light of your estimates from this problem and the previous one,
what do you think is going on with these foxes?


### Estimand 

What is the direct causal influence of average food on weight? 


### Scientific model

```{r}
#| echo: false
coords <- data.frame(
	name = c('F', 'G', 'A', 'W'),
	x =    c(1,    2,   1.5, 1.5),
	y =    c(0,    0,   1, -1)
)
```

```{r}
dag <- dagify(
	G ~ F,
	F ~ A,
	W ~ F + G,
  coords = coords,
	exposure = 'F',
	outcome = 'W'
)
ggdag_status(dag, seed = 2, layout = 'auto') + theme_dag()
```


#### Backdoor criterion

1. Identify all paths connecting treatment to the outcome, regardless of the
direction of arrows

- `F -> W`
- `F -> G -> W`

2. Identify paths with arrows entering the treatment (backdoor). These are 
non-casual paths, because causal paths exit the treatment (frontdoor). 

- ~~`F -> W`~~
- ~~`F -> G -> W`~~


3. Find adjustment sets that close all backdoor/non-causal paths. 

There are no backdoor paths entering the treatment (F). However, unlike in Q02, 
we are interested in the direct effect. Therefore, the adjustment set includes
group size. 

```{r}
ggdag_adjustment_set(dag, effect = 'direct') + theme_dag()
```


### Statistical model

$$W_{i} \sim Normal(\mu_{i}, \sigma)$$
$$\mu_{i} = \alpha + \beta_{F} F_{i} + \beta_{G} G_{i}$$

$$\alpha \sim Normal(0, 0.2)$$

$$\beta_{G} \sim Normal(0, 0.5)$$

$$\beta_{F} \sim Normal(0, 0.5)$$

$$\sigma \sim Exponential(1)$$

### Analyze the data

```{r}
# Print priors used
tar_read(h03_q03_brms_prior)

# Load model
tar_load(h03_q03_brms_sample)
h03_q03_brms_sample

# Tidy draws
q03_draws <- h03_q03_brms_sample |> 
    tidy_draws()

# New data, with three levels of group size (L, M, H)
q03_newdata <- data_grid(DT, scale_avgfood = seq_range(scale_avgfood, 20),
												 scale_groupsize = seq_range(scale_groupsize, 3))
q03_newdata$group_size <- as.factor(q03_newdata$scale_groupsize)
levels(q03_newdata$group_size) <- c('L', 'M', 'H')

q03_epred <- h03_q03_brms_sample |> 
    epred_draws(q03_newdata)

# Direct causal influence of area on avg food
g1 <- ggplot(q03_epred) +
	stat_ribbon(aes(
		reverse_scale(scale_avgfood, attributes(DT$scale_avgfood)),
		reverse_scale(.epred, attributes(DT$scale_weight))
	), alpha = 0.5) +
	scale_fill_grey(start = 0.8, end = 0.2) +
	labs(
		x = 'Average food',
		y = 'Weight',
		title = paste(
			'Direct causal influence of average food on weight,',
			'\n',
			'at low, medium and high group size'
		)
	) +
	facet_wrap( ~ group_size)

g2 <- ggplot(q03_draws) +
	geom_vline(xintercept = 0, color = 'grey70') +
	stat_halfeye(aes(b_scale_avgfood)) +
	labs(x = TeX(r"($\beta_{F}$)"), y = '') + 
	xlim(-1.2, 1.2)

g1 / g2
```


Holding group size constant, we see a positive association between average 
food and weight. We can expect that fox in larger groups have access to
more available food, they also have to share with more conspecifics. 


## Question 4

> Suppose there is an unobserved confound that influences F and G, like [below].
Assuming the DAG above is correct, again estimate both the total and direct
causal effects of F on W. What impact does the unobserved confound have?


### Estimand 

What is the direct and total effect of F on W? What impact does the unobserved
confound U have?


### Scientific model

```{r}
#| echo: false
coords <- data.frame(
	name = c('F', 'G', 'A', 'W', 'U'),
	x =    c(1,    2,   1, 1.5,   2),
	y =    c(0,    0,   1, -1,    1)
)
```

```{r}
dag <- dagify(
	G ~ F + U,
	F ~ A + U,
	W ~ F + G,
  coords = coords,
	exposure = 'F',
	outcome = 'W',
	latent = 'U'
)
ggdag_status(dag, seed = 2, layout = 'auto') + theme_dag()
```


#### Backdoor criterion

1. Identify all paths connecting treatment to the outcome, regardless of the
direction of arrows

- `F -> W`
- `F -> G -> W`
- `F <- U -> G -> W`

2. Identify paths with arrows entering the treatment (backdoor). These are 
non-casual paths, because causal paths exit the treatment (frontdoor). 

- ~~`F -> W`~~
- ~~`F -> G -> W`~~
- `F <- U -> G -> W`


3. Find adjustment sets that close all backdoor/non-causal paths. 

There is one backdoor path entering the treatment (F) through the 
unobserved confound U. The total causal effect cannot be estimate 
because of the unobserved confound introducing a backdoor path into 
the treatment that cannot be controlled for. The direct causal effect
can be estimated as in Q03 by including G in the adjustment set. 

```{r}
ggdag_adjustment_set(dag, effect = 'total')  + theme_dag()
ggdag_adjustment_set(dag, effect = 'direct') + theme_dag()
```


