---
title: "Homework 04"
author: "Alec L. Robitaille"
date: "2024-03-13"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> Revisit the marriage, age, and happiness collider bias example from Chapter
6. Run models m6.9 and m6.10 again (pages 178–179). Compare these two models
using both PSIS and WAIC. Which model is expected to make better predictions,
according to these criteria, and which model yields the correct causal
inference?


```{r}

```


## Question 2

> Reconsider the urban fox analysis from last week’s homework. On the basis
of PSIS and WAIC scores, which combination of variables best predicts body
weight (W, weight)? What causal interpretation can you assign each coefficient
(parameter) from the best scoring model?



## Question 3 

> The data in data(Dinosaurs) are body mass estimates at different estimated
ages for six different dinosaur species. See ?Dinosaurs for more details. Choose
one or more of these species (at least one, but as many as you like) and model
its growth. To be precise: Make a predictive model of body mass using age as a
predictor. Consider two or more model types for the function relating age to
body mass and score each using PSIS and WAIC. Which model do you think is best,
on predictive grounds? On scientific grounds? If your answers to these questions
differ, why? This is a challenging exercise, because the data are so scarce. But
it is also a realistic example, because people publish Nature papers with even
less data. So do your best, and I look forward to seeing your growth curves.
