---
title: "Homework 05"
author: "Alec L. Robitaille"
date: "2024-03-28"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> The data in data(NWOGrants) are outcomes for scientific funding applications
for the Netherlands Organization for Scientific Research (NWO) from 2010–2012
(see van der Lee and Ellemers doi:10.1073/pnas.1510159112). These data have a
structure similar to the UCBAdmit data discussed in Chapter 11 and in lecture.
There are applications and each has an associated gender (of the lead
researcher). But instead of departments, there are disciplines. Draw a DAG for
this sample. Then use the backdoor criterion and a binomial GLM to estimate the
TOTAL causal effect of gender on grant awards.

### Estimand 

What is the total causal effect of gender on grant awards?

### Scientific model

```{r}
#| echo: false
coords <- data.frame(
	name = c('G', 'D', 'A'),
	x =    c(1,    2,   3),
	y =    c(0,    1,   0)
)
```

```{r}
dag <- dagify(
	D ~ G, 
	A ~ G + D,
  coords = coords,
	exposure = 'G',
	outcome = 'A'
)
ggdag_status(dag, seed = 2, layout = 'auto') + theme_dag()
```


#### Backdoor criterion

1. Identify all paths connecting treatment to the outcome, regardless of the
direction of arrows

- `G -> A`
- `G -> D -> A`

2. Identify paths with arrows entering the treatment (backdoor). These are 
non-casual paths, because causal paths exit the treatment (frontdoor). 

- ~~`G -> A`~~
- ~~`G -> D -> A`~~


3. Find adjustment sets that close all backdoor/non-causal paths. 

There are no backdoor paths entering the treatment (G). There 
is a direct path from G -> A and an indirect path through D. 
The adjustment set for the total effect is empty. 

```{r}
ggdag_adjustment_set(dag, effect = 'total') + theme_dag()
```


### Statistical model

$$A_{i} \sim Binomial(p_{i})$$
$$p_{i} = \alpha + \beta_{G} G_{i}$$
$$\alpha \sim Normal(0, 0.2)$$
$$\beta_{G} \sim Normal(0, 0.5)$$
$$\sigma \sim Exponential(1)$$

#### Prior predictive simulation

```{r}
```


### Analyze the data

```{r}
data_grants
```

```{r}

```


## Question 2

> Now estimate the DIRECT causal effect of gender on grant awards. Use the same
DAG as above to justify one or more binomial models. Compute the average direct
causal effect of gender, weighting each discipline in proportion to the number
of applications in the sample. Refer to the marginal effect example in Lecture 9
for help


```{r}
```

## Question 3 

> OPTIONAL CHALLENGE. The data in data(UFClefties) are the outcomes of 205
Ultimate Fighting Championship (UFC) matches (see ?UFClefties for details). It
is widely believed that left-handed fighters (aka “Southpaws”) have an advantage
against right-handed fighters, and left-handed men are indeed over-represented
among fighters (and fencers and tennis players) compared to the general
population. Estimate the average advantage, if any, that a left-handed fighter
has against right-handed fighters. Based upon your estimate, why do you think
left-handers are over-represented among UFC fighters?

```{r}
```
