---
title: "Homework 06"
author: "Alec L. Robitaille"
date: "2024-04-11"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> Conduct a prior predictive simulation for the Reedfrog model. By this I mean
to simulate the prior distribution of tank survival probabilities. Start by
using this prior:

$$\alpha_{j} \sim Normal(\bar{\alpha}, \sigma)$$
$$\bar{alpha} \sim Normal(0, 1)$$
$$\sigma \sim Exponential(1)$$

> Be sure to transform the alpha j values to the probability scale for plotting
and summary. How does increasing the width of the prior on sigma change the
prior distribution of alpha j? You might try Exponential(10) and
Exponential(0.1) for example.


#### Prior predictive simulation

```{r}
# Load data
tar_load(reedfrog)

# Print priors used
tar_read(h06_q01_exp_1_brms_prior)
tar_read(h06_q01_exp_0pt1_brms_prior)
tar_read(h06_q01_exp_10_brms_prior)

# Load model
tar_load(h06_q01_exp_1_brms_sample_prior)
tar_load(h06_q01_exp_0pt1_brms_sample_prior)
tar_load(h06_q01_exp_10_brms_sample_prior)

# Read N draws from the priors and append expected predictions
n_draws <- 100
q01_exp_1_prior <- h06_q01_exp_1_brms_sample_prior |>
	add_predicted_draws(newdata = reedfrog, ndraws = n_draws)
q01_exp_0pt1_prior <- h06_q01_exp_0pt1_brms_sample_prior |>
	add_predicted_draws(newdata = reedfrog, ndraws = n_draws)
q01_exp_10_prior <- h06_q01_exp_10_brms_sample_prior |>
	add_predicted_draws(newdata = reedfrog, ndraws = n_draws)


# Plot prior expectations for survival rate and tank
g1 <- ggplot(q01_pred_prior) + 
	stat_halfeye(aes(tank, .prediction), alpha = 0.5) +
	labs(x = 'Tank', y = 'Predicted survival rate') + 
	xlim(0, 1)
g0pt1 <- ggplot(q01_pred_prior) + 
	stat_halfeye(aes(tank, .prediction), alpha = 0.5) +
	labs(x = 'Tank', y = 'Predicted survival rate') + 
	xlim(0, 1)
g10 <- ggplot(q01_pred_prior) + 
	stat_halfeye(aes(tank, .prediction), alpha = 0.5) +
	labs(x = 'Tank', y = 'Predicted survival rate') + 
	xlim(0, 1)

g1 / g0pt1 / g10
```


## Question 2

> Revisit the Reedfrog survival data, data(reedfrogs). Start with the varying
effects model from the book and lecture. Then modify it to estimate the causal
effects of the treatment variables pred and size, including how size might
modify the effect of predation. An easy approach is to estimate an effect for
each combination of pred and size. Justify your model with a DAG of this
experiment

### Estimand
