---
title: "Homework 06"
author: "Alec L. Robitaille"
date: "2024-04-11"
---

## Setup

```{r}
#| include: false
targets::tar_source('R')
```


```{r}
#| eval: false
targets::tar_source('R')
```


## Question 1

> Conduct a prior predictive simulation for the Reedfrog model. By this I mean
to simulate the prior distribution of tank survival probabilities. Start by
using this prior:

$$\alpha_{j} \sim Normal(\bar{\alpha}, \sigma)$$
$$\bar{\alpha} \sim Normal(0, 1)$$
$$\sigma \sim Exponential(1)$$

> Be sure to transform the alpha j values to the probability scale for plotting
and summary. How does increasing the width of the prior on sigma change the
prior distribution of alpha j? You might try Exponential(10) and
Exponential(0.1) for example.


#### Prior predictive simulation

```{r}
# Load data
tar_load(reedfrogs)
```

::: {.panel-tabset}

## Exponential(1)
```{r}
tar_load(h06_q01_exp_1_brms_sample_prior)
h06_q01_exp_1_brms_sample_prior$prior
```


```{r prior_exp_1_areas}
#| fig-width: 7
#| fig-height: 12
mcmc_areas(
	h06_q01_exp_1_brms_sample_prior,
	regex_pars = 'r_tank', 
	transformations = inv_logit
)
```

## Exponential(0.1)

```{r}
tar_load(h06_q01_exp_0pt1_brms_sample_prior)
h06_q01_exp_0pt1_brms_sample_prior$prior
```


```{r prior_exp_0pt1_areas}
#| fig-width: 7
#| fig-height: 12
mcmc_areas(
	h06_q01_exp_0pt1_brms_sample_prior,
	regex_pars = 'r_tank',
	transformations = inv_logit
)
```

## Exponential(10)

```{r}
tar_load(h06_q01_exp_10_brms_sample_prior)
h06_q01_exp_10_brms_sample_prior$prior
```


```{r prior_exp_10_areas}
#| fig-width: 7
#| fig-height: 12
mcmc_areas(
	h06_q01_exp_10_brms_sample_prior, 
	regex_pars = 'r_tank', 
	transformations = inv_logit
)
```
:::




## Question 2

> Revisit the Reedfrog survival data, data(reedfrogs). Start with the varying
effects model from the book and lecture. Then modify it to estimate the causal
effects of the treatment variables pred and size, including how size might
modify the effect of predation. An easy approach is to estimate an effect for
each combination of pred and size. Justify your model with a DAG of this
experiment

### Estimand

What is the causal effect of predation and size on survival?

### Scientific model

```{r}
coords <- data.frame(
	name = c('D', 'G', 'P', 'T', 'S'),
	x =    c(1,    2,   3,   1,    2),
	y =    c(0,    0,   0,   1,  1)
)
```

```{r}
#| fig.height: 4
#| fig.width: 4 
dag <- dagify(
	S ~ D + G + P + T,
  coords = coords,
	outcome = 'S',
	exposure = c('G', 'P')
)
ggdag(dag, seed = 2, layout = 'auto') + theme_dag()
```

- 48 tanks (T) of reedfrogs
- treatments: density (D), size (G), predation (P)
- outcome: survival (S)


### Adjustment set

```{r}
ggdag_adjustment_set(dag, effect = 'direct') + theme_dag()
```


### Analyze the data

```{r}
# Load model
tar_load(h06_q02_brms_sample)
h06_q02_brms_sample

# Estimated marginal effects
marg_eff <- emmeans(h06_q02_brms_sample, ~ size*pred, regrid = 'response')
marg_eff
plot(marg_eff) + labs(x = 'Survival', y = 'Size : Predation')
```



## Question 3

> 3-OPTIONAL CHALLENGE. Return to the Trolley data, data(Trolley), from Chapter
12. Define and fit a varying intercepts model for these data. By this I mean to
add an intercept parameter for the individual participants to the linear model.
Cluster the varying intercepts on individual participants, as indicated by the
unique values in the id variable. Include action, intention, and contact as
treatment effects of interest. Compare the varying intercepts model and a model
that ignores individuals. What is the impact of individual variation in these
data?

### Estimand

What is the causal effect of action, intention and contact on survey 
responses, considering the variation of individuals?

### Scientific model

```{r}
coords <- data.frame(
    name = c('X', 'R', 'S', 'E', 'Y', 'G',  'P'),
    x =    c(1,    2,   3,   1,    2,   3,  2),
    y =    c(0,    0,   0,   -2,  -1,  -2,  -3)
)

dag <- dagify(
    R ~ X + S + E + Y + G,
    E ~ Y + G,
  coords = coords
)
ggdag(dag, seed = 2, layout = 'auto') + theme_dag()
```

- R: response to the trolley story
- X: action, intention, contact
- S: story
- E: education
- Y: age
- G: gender


### Analyze the data

```{r}
tar_load(trolley)

tar_load(h06_q03_no_ind_var_brms_sample)
tar_load(h06_q03_ind_var_brms_sample)

h06_q03_no_ind_var_brms_sample
h06_q03_ind_var_brms_sample
```
