---
title: "Lecture 03 Notes"
author: "Alec L. Robitaille"
date: "2024-02-08"
---


```{r}
#| include: false
source('R/packages.R')
```

# Geocentric models

Prediction without explanation. The geocentric model has epicycles and Earth 
is the center of the universe. 

Linear regressions are geocentric - they describe associations, make predictions
but are mechanically wrong. Therefore, they are useful when handled with care. 


## Normal distribution

Summed fluctuations tend towards the normal distribution

For estimating only the mean and variance, the normal distribution is the 
least informative distribution (maxent).

Note: variables do not have to be normally distributed for normal models 
to be useful. 


# Linear regression

## Workflow

1. Question/goal/estimand
1. Scientific model
1. Statistical model(s)
1. Validate model
1. Analyze data


## Example: Howell data

### 1. Estimand

Describe the association between adult weight and height

### 2. Scientific model

How does height influence weight? 

H -> W <- U

```{r}
coords <- data.frame(
	name = c('H', 'W', 'U'),
	x =    c(1,    2,   3),
	y =    c(0,    0,  0)
)
```

```{r}
#| fig-height: 4
dagify(
	W ~ H + U,
  coords = coords
) |> ggdag(seed = 2, layout = 'auto') + theme_dag()
```

This DAG corresponds to: "weight is some function of height and unobserved variables" ($W = f(H, U)$).


$$W = \beta H + U$$

The unobserved variable U is normally distribution with a mean of 0 and 
a standard deviation sigma.

$$U_{i} \sim Normal(0, \sigma)$$

Height is uniformly distributed between 130 and 170 cm. 

$$H_{i} \sim Uniform(130, 170)$$


There is a deterministic relationship for W and distributional relationships
for U and H. 



Corresponding generative code is:

```{r}
sim_weight <- function(H, beta, sigma) {
	U <- rnorm(length(H), mean = 0, sd = sigma)
	W <- beta * H + U
	return(W)
}

heights <- seq.int(10, 50)
sim_weight(heights, beta = 0.5, sigma = 0.1)
```

### 3. Statistical model

$E(W_{i} | H_{i}) = \alpha + \beta H_{i}$

- $E(W_{i} | H_{i})$: average weight conditional on height
- $\alpha$: intercept
- $\beta H_{i}$: slope


#### Posterior distribution

$Pr(\alpha, \beta, \sigma | H_{i}, W_{i}) = \frac{Pr(W_{i} | H_{i}, \alpha, \beta, \sigma) Pr (\alpha, \beta, \sigma)}{Z}$


- $Pr(\alpha, \beta, \sigma | H_{i}, W_{i})$: posterior probability of a specific line
- $Pr(W_{i} | H_{i}, \alpha, \beta, \sigma)$: garden of forking data, the number of ways 
we could see the observations, conditional on an exact line (in this case)
- $Pr (\alpha, \beta, \sigma)$: prior
- $Z$: normalizing constant

Alpha, beta and sigma are unobserved variables, height and weight are observed. 

The posterior distribution is proportional to the product of the number of ways
the observations could arise according to our assumptions multiplied by our
prior. 


#### Prior predictive simulation

What do the observable variables look like with these priors?

$W_{i} \sim Normal(\mu_{i}, \sigma)$

$\mu_{i} = \alpha + \beta H_{i}$

$\alpha \sim Normal(0, 10)$

$\beta \sim Uniform(0, 1)$

$\sigma \sim Uniform(0, 10)$


When height is 0, weight is 0. Weight increases on average with height. Weight
(kg) is less than height (cm). Sigma must be positive.


Priors: 

- no correct priors, only scientifically justifiable priors. 
- justify with information outside the data, like the rest of the model
- priors become more important as models get more complex
- always simulate from the priors, better than trying to intuit what they mean
from the definitions


### 4. Validate model

Test statistical model with simulated observations from scientific model. 
Stronger test: simulation-based calibration

- simulate individuals with known parameters from scientific model
- use model to determine if it can recover parameters 
- use a large sample
- test with different values and sample sizes


### 5. Analyze the data

Once you've done steps 1-4, you are ready to analyze the real data.

Parameters are not independent, do not interpret them separately. 

Use the posterior distribution to generate predictions on the outcome 
scale then describe and interpret those. 

