---
title: "Lecture 13 Notes"
author: "Alec L. Robitaille"
date: "2024-04-17"
---

```{r, include = FALSE}
source('R/packages.R')
```


# Clusters and features

Clusters: kinds of groups in the data

Features: aspects of the model (parameters) that vary by cluster

Examples (clusters -> features): 

- tanks -> survival
- stories -> treatment effect
- individuals -> average responses
- departments -> admission rate, bias


Adding more clusters involves adding more index variables and more population
priors.

Adding more features involves adding more parameters, increasing the dimensions
in each population prior.


Varying effects are a way to measure unmeasured confounds. Unmeasured features
of clusters leave an imprint on the data that can be estimated given
repeat observations of each cluster and partial pooling among clusters. 

This helps us, from the predictive perspective, to regularize cluster-level 
variance, and, from the causal perspective, to manage unobserved confounds
and competing causes using cluster-level causes. 

Practical difficulties of varying effects:

- How to use more than one cluster type at the same time?
- How to calculate predictions? At which level?
- How to sample chains efficiently?
- Group-level confounding


# Strategy for modeling varying effects 

1. Varying intercept on one cluster
1. Varying intercept for one cluster, slope for another cluster. This
is two 1-dimensional distributions where information is not shared across. 
1. Correlated varying effects


## Example: Bangladesh fertility survey

Outcome: contraceptive use

Variables: age, living children, urban/rural, districts

```{r}
#| fig.height: 4
#| fig.width: 4 

coords <- data.frame(
	name = c('A', 'C', 'D', 'K', 'U'),
	x =    c(1,    2,    3,   1.5, 2.5),
	y =    c(0,    0.5,    0,  -1, -1)
)
dagify(
	C ~ A + K + D + U,
	K ~ A + U,
	U ~ D,
  coords = coords
) |> ggdag(seed = 2, layout = 'auto') + theme_dag()
```



In addition,

- group level confounds: unmeasured things about districts may influence features
of individuals in those districts (eg. history of districts)

- unmeasured variables: family
