---
title: "Lecture 16 Notes"
author: "Alec L. Robitaille"
date: "2024-05-22"
---

```{r, include = FALSE}
source('R/packages.R')
```


# Gaussian processes

Gaussian process is "an infinite-dimensional generalization
of multivariate normal distributions". Instead of a conventional 
covariance matrix, use a kernel function that generalizes to infinite
dimensions/observations/predictions. It allows you to use a smaller 
number of parameters, inside the covariance matrix, with regularization. 

Given the kernel function, the covariance can get arbitrarily large 
because it doesn't add any more parameters. 

The kernel function can be based on differences, eg. space, time, age. 
These are continuous ordered categories. Partial pooling where points 
that are closer together to pool more. 

The kernel function describes the expected covariance between any
two points separated by a given distance. The kernel function is estimated 
along with the other parameters. 

Also see:

- Automatic relevance determination (ARD) for multiple distance dimensions
- Multi-output Gaussian processes, eg. multiple traits at the same time
with covariation
- Telemetry, navigation (Kalman filter)


## Kernel functions

Goal is to describe the macro state, "what's the shape of decline
in covariation". Some options:

Quadratic (L2)

- Gaussian covariance kernel

Ornstein-Uhlenbeck

- exponential decline

Periodic

- cyclical variables eg. time of day, orientation

## Example: oceanic technology

Number of tool types associated with population size

Spatial covariation where islands close together share unobserved 
confounds and innovations. The effect of spatial covariation is to make
closer islands have more similar tools or change in tools. 

```{r}
#| fig.height: 4
#| fig.width: 5
coords <- data.frame(
	name = c('T', 'C', 'P', 'U'),
	x =    c(1,    2,   0,  1),
	y =    c(1,    1,   1,  0)
)
dagify(
	T ~ P + C + U,
	P ~ U,
  coords = coords
) |> ggdag(seed = 2, layout = 'auto') + theme_dag()
```

- T: tools
- P: population
- C: contact
- U: unobserved confounds

### Statistical model without spatial covariation

Functional relationship

$$\Delta(T) = \alpha P^{\beta} - \gamma T$$

- $\Delta(T)$: change in tools
- $\alpha$: innovation rate
- $\beta$: diminishing returns
- $\gamma$: rate of loss

Statistical model

$$T_{i} \sim Poisson (\lambda_{i})$$
$$\lambda_{i} = \hat{T}$$
$$\hat{T} = \frac{\alpha P ^{\beta}}{\gamma}$$


### Statistical model with spatial covariation, but without population

Varying intercepts model


$$T_{i} \sim Poisson (\lambda_{i})$$

$$log \lambda_{i} = \bar{\alpha} + \alpha_{S_{[i]}}$$
$$\hat{T} = \frac{\alpha P ^{\beta}}{\gamma}$$ 

- $\bar{\alpha}$: mean across islands
- $\alpha_{S_{[i]}$: deviation from mean by island

$$\begin{bmatrix} \alpha_{1} \\ \alpha_{2} \\ .. \\ \alpha_{10} \end{bmatrix} \sim MVNormal(\begin{bmatrix} 0 \\ 0 \\ ... \\ 0 \end{bmatrix}, K)$$
- vector of varying effects
- vector of zeros
- covariance matrix, a 10x10 kernel


